\documentclass[acmlarge]{acmart}
\usepackage[utf8]{inputenc}

\usepackage{booktabs} % For formal tables
\usepackage{tipa}
\newcommand{\tttt}[1]{\texttt{#1}}
\usepackage{subcaption}


\usepackage[ruled]{algorithm2e} % For algorithms
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

% Metadata Information
\acmJournal{TALLIP}
% \acmVolume{X}
% \acmNumber{X}
% \acmArticle{XX}
% \acmYear{2018}
% \acmMonth{X}
% \acmArticleSeq{XX}
\acmVolume{0}
\acmNumber{0}
\acmArticle{00}
\acmYear{2018}
\acmMonth{7}
\acmArticleSeq{00}


%\acmBadgeR[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
%\acmBadgeL[http://ctuning.org/ae/ppopp2016.html]{ae-logo}


% Copyright
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
\acmDOI{0000001.0000001}

% Paper history
\received{July 2018}
% \received{March 2009}
\received[accepted]{XXX XXXX}


% Document starts
\begin{document}
% Title portion
\title{Basic concepts and tools for the Toki Pona minimalist and
constructed language}
\titlenote{Description of the language and main issues;
Wordnet synsets; analysis of the vocabulary; synthesis and syntax highlighting of texts}

\author{Renato Fabbri}
\authornote{This is the corresponding author}
\orcid{0000-0002-9699-629X}
\affiliation{%
  \institution{Institute of Mathematical and Computer Sciences, University of São Paulo}
  \streetaddress{Avenida Trabalhador São-carlense, 400 - Centro}
  \city{São Carlos}
  \state{SP}
  \postcode{668}
  \country{BR}}
\email{renato.fabbri@gmail.com}



\begin{abstract}
A minimalist constructed language (conlang)
is useful for experiments and comfortable for making tools.
The Toki Pona (TP) conlang is minimalist both in the vocabulary
(with only 14 letters and 124 words)
  and in the (about) ten syntax rules.
The language is useful for being a used and somewhat established
minimalist conlang with at least hundreds of fluent speakers.
In this article, we describe current concepts and resources
for TP,
and make available Python scripted routines for
the analysis of the language,
the synthesis of texts, the specification of syntax highlighting
schemes, and the achievement of a preliminary TP
Wordnet.
We focus on the analysis of the basic vocabulary,
as corpus analyses were found.
The synthesis is based on sentence templates,
relates to context by keeping track of used words,
and renders larger texts by using a fixed number of phonemes (e.g. for poems)
and number of sentences, words and letters (e.g. for paragraphs).
Syntax highlighting 
reflects morphosyntactic classes given in the official dictionary
and different solutions are described and implemented
in the well-established Vim text editor.
The tentative TP Wordnet is made available in three forms that
reflect the choices of the synsets related to each word.
In summary, this text holds potentially novel conceptualizations about,
and tools and results in analyzing, synthesizing and syntax highlighting the
TP language.
\end{abstract}


%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below.
%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010405.10010469</concept_id>
  <concept_desc>Applied computing~Arts and humanities</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010405.10010497</concept_id>
  <concept_desc>Applied computing~Document management and text processing</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10003456.10010927.10003619</concept_id>
  <concept_desc>Social and professional topics~Cultural characteristics</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc{Applied computing~Arts and humanities}
\ccsdesc[500]{Applied computing~Document management and text processing}
\ccsdesc[300]{Social and professional topics~Cultural characteristics}

%
% End generated code
%


\keywords{Constructed languages, natural language processing, syntax highlighting, wordnet, Toki Pona}

% DO NOT use this command unless you want to change
% the default behavior
% \authorsaddresses{Authors' addresses: G.~Zhou, Computer Science
%   Department, College of William and Mary, 104 Jameson Rd,
%   Williamsburg, PA 23185, US, \path{gzhou@wm.edu}; V.~B\'eranger,
%   Inria Paris-Rocquencourt, Rocquencourt, France; A.~Patel, Rajiv
%   Gandhi University, Rono-Hills, Doimukh, Arunachal Pradesh, India;
%   H.~Chan, Tsinghua University, 30 Shuangqing Rd, Haidian Qu, Beijing
%   Shi, China; T.~Yan, Eaton Innovation Center, Prague, Czech Republic;
%   T.~He, C.~Huang, J.~A.~Stankovic University of Virginia, School of
%   Engineering Charlottesville, VA 22903, USA; T. F. Abdelzaher,
%   (Current address) NASA Ames Research Center, Moffett Field,
%   California 94035.}

\maketitle

% The default list of authors is too long for headers.
\renewcommand{\shortauthors}{G. Zhou et al.}

\input{samplebody-journals}



\end{document}
